Group 10 Final Project

### 2024-08-06 manuscript_final-2.qmd
uypdate final

#remove.packages("here")

@@

Group Ten is conducting a comprehensive analysis of historical data from
various public agencies to evaluate the impact of Varroa mites, American
Foulbrood, and global warming on hive losses in the United States.This
research utilizes extensive datasets from the National Agricultural
Statistics Service, the Agricultural Statistics Board, and the United
States Department of Agriculture (USDA), encompassing multiple years of
data. The data highlights the topic on hive losses attributed to mites,
bacterial infections, and environmental factors related to global
warming. Through advanced data visualization techniques in R, we aim to
demonstrate and validate the detrimental effects of these factors on
honeybee colonies, highlighting the consequent implications for honey
production and broader food security. ***Dennis and Kemp's study on
honey bee hive collapse [@hivecollapse] provides important insights into
Allee effects and ecological resilience.***

@@

Honeybees have served as nature's pollinators for centuries. With their
relationship with humans documented as far back as ancient Egyptian and
Hindu cultures. Historically, humans have maintained beehives and
utilized honey for its medicinal properties in various civilizations,
including the Egyptians, Assyrians, Chinese, Greeks, and Romans. Honey's
natural antibacterial qualities made it a valuable treatment for wounds.
This was a practice continued by Romans and Russians during World War I.
Honeybees and other pollinators are essential for food production and
nutritional security. But even though they face numerous survival
challenges Varroa mites pose a significant threat to bee colonies. These
tiny red-brown parasites live on adult honeybees and reproduce on larvae
and pupae. Another major threat is American Foulbrood Disease (AFB).
This is caused by the bacterium Paenibacillus larvae, which is fatal to
honeybee larvae and requires incineration of infected hives to prevent
its spread. The impact of climate change on honeybee colony losses is a
growing area of research. While correlations have been found between
higher winter temperatures and increased colony losses, the effects of
warmer autumn and winter temperatures on colony dynamics and survival
require further investigation. ***Dennis and Kemp's study on honey bee
hive collapse [@hivecollapse] provides important insights into Allee
effects and ecological resilience.***[@usda_honey_bees; @usda_nass]

@@

# Plot with ggplot2
ggplot(honey_data, aes(x = "", y = percentage, fill = component)) +
  geom_bar(width = 0.8, stat = "identity", color = "white") +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = honey_data$color) +
  geom_text(aes(x=1.8, label = paste0(percentage, "%")), position = position_stack(vjust = 0.6), size = 3, color = "black") +
  labs(
    title = "COMPOSITION OF HONEY",
    subtitle = "Illustration of honey components by percentage",
    fill = NULL
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5, family = "lobster"),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 12)
  )
```
### Composition of Honey

Compositions of honey is a pie chart that illustrates the composition of honey
by percentage. The main components are:

@@

```
### Spread of Varroa Mites

the spread of varroa mites image is a map illustrating the spread of Varroa mites in the
United States from 1986 to 1995. The map uses different colors to
represent the year of the first appearance of Varroa mites in each
state:

@@

The map reveals the progressive spread of Varroa mites from the
mid-1980s to the mid-1990s, affecting bee colonies across the country.
The earliest appearances were concentrated in specific regions and
gradually spread to more states over time. This visual representation
highlights the increasing geographical distribution of Varroa mites and
underscores the growing challenge they pose to beekeeping and
agricultural industries in the United States.

***[@varroamap] Varroa Map of infestation periods***

@@

### Detailed Description:
- **Trend Analysis**:
  - From 1900 to around 1975, the Earth's surface temperature shows considerable variability with some notable periods of cooling and warming.
  - Post-1975, there is a distinct upward trend in temperature anomalies for all three measurements, indicating a general warming trend.
- **Measurement Comparison**:
  - The Earth's surface (blue line) shows the most extended historical data and fluctuates more in the earlier years compared to the troposphere measurements.
  - The lower troposphere measurements (RSS in red and UAH in green) start around 1975 and generally follow the same upward trend as the surface measurements but with less variability.
  - The RSS and UAH measurements align closely with each other post-2000, indicating consistent trends in the lower troposphere temperature anomalies.
- **Anomalies & Peaks**:
  - The Earth's surface shows significant peaks and troughs, especially in the early 20th century and around the year 2000.
  - The lower troposphere measurements show peaks and align closely, particularly after the year 2000, reflecting similar patterns in temperature anomalies.

### Key Points:
- The overall trend shows an increase in temperature anomalies from 1900 to 2025.
- Significant variability is seen in Earth's surface temperatures in the early years.
- Post-1975, a clear warming trend is observed across all measurements.
- Lower troposphere data (RSS and UAH) begins around 1975 and follows the Earth's surface trend with less variability.
- Post-2000, the RSS and UAH data closely align, indicating consistent temperature trends in the lower troposphere.
- There are noticeable peaks in the temperature anomalies around the year 2000 and later, reflecting periods of significant temperature increases.

### Impact on Bees, Varroa Mites, and Colony Loss:
- **Bees**:
  - **Thermal Stress**: Increased temperature anomalies can cause thermal stress on bees, affecting their ability to forage, navigate, and perform essential colony tasks.
  - **Floral Resources**: Warming temperatures can alter the availability and distribution of floral resources, impacting bee nutrition and health.
  - **Reproduction Cycles**: Changes in temperature can disrupt the reproductive cycles of bees, potentially leading to mismatches with the availability of pollen and nectar.

- **Varroa Mites**:
  - **Increased Reproduction**: Warmer temperatures can accelerate the reproduction rates of varroa mites, exacerbating infestations within bee colonies.
  - **Extended Activity Period**: Higher temperatures can lengthen the active period of varroa mites throughout the year, leading to prolonged periods of stress on bee populations.

- **Colony Loss**:
  - **Health Decline**: Combined effects of thermal stress on bees and increased varroa mite infestations can lead to a decline in colony health.
  - **Colony Collapse Disorder (CCD)**: Temperature anomalies can contribute to conditions that favor Colony Collapse Disorder, where the majority of worker bees disappear, leaving behind the queen and immature bees.
  - **Mortality Rates**: Sustained temperature increases can lead to higher mortality rates within bee colonies, significantly impacting beekeeping and agricultural pollination services.

### Conclusion:
The chart highlights a clear trend of increasing temperature anomalies, which can have profound implications on bee populations, varroa mite dynamics, and overall colony health. Understanding these trends is crucial for developing strategies to mitigate the adverse effects on bees and ensure the sustainability of pollination services essential for ecosystems and agriculture.

[@epa_climate_indicator] ***Global warming trends from 1900-2024***

@@

1. **USDA National Agricultural Statistics Service (NASS)**:
   - Provides comprehensive agricultural data, including statistics on honey production and colony health.
   - Source: [USDA - National Agricultural Statistics Service - Surveys - honey bee surveys and reports](https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/).

2. **Bee Informed Partnership**:
   - Offers detailed insights and research on bee colony management and health, contributing valuable information on the status and trends of bee populations.
   - Source: [Index Catalog // USDA Economics, Statistics and Market Information System](https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en).

@@

### Hypotheses

@@

#Citation section To cite other work (important everywhere, but likely
happens first in introduction), make sure your references are in the
bibtex file specified in the YAML header above and have the right bibtex
key. Then you can include like this:

Examples of reproducible research projects can for instance be found in
[@mckay2020; @mckay2020a].

@@

Using this in RMarkdown: #Make sure to save these entries in a file
names'reference.bib' and include it in your R Markdown document like
this:

@@

# Install necessary packages
if (!require("DiagrammeR")) install.packages("DiagrammeR")

# Create the flowchart using DiagrammeR
grViz("
  digraph Data_Analysis_Flowchart {
    graph [layout = dot, rankdir = LR, label = 'Workflow Schematic', fontsize = 45, labelloc = 't']

    node [style = filled, fontsize = 30, width = 2, height = 2]

    Define_Objectives -> Collect_Data
    Collect_Data -> Clean_Data
    Clean_Data -> Analyze_Data
    Analyze_Data -> Interpret_Results
    Interpret_Results -> Visualize_Data
    Visualize_Data -> Compile_Report
    Compile_Report -> Review_Revise
    Review_Revise -> Present_Report
  }
")
```

#knitr::include_graphics(here("assets","antigen-recognition.png"))
```

@@

### Data Acquisition

We sourced our data from the United States Department of Agriculture (USDA), leveraging their comprehensive and authoritative datasets. This data provided crucial information for our analysis, including statistics on honey production, colony health, and factors affecting bee populations. Utilizing USDA data ensured the reliability and relevance of our analysis, as it encompasses extensive, well-documented records pertinent to agricultural and ecological studies.

@@

1. **Removal of Irrelevant Data**:
   - **Blank Spaces**: We eliminated any blank spaces within the datasets to ensure data integrity.
   - **Non-Pertinent Columns**: Columns that were not directly related to our analysis objectives were removed to streamline the dataset.

2. **Filtering Observations**:
   - **Focus on Relevant Data**: We filtered out observations that did not directly pertain to our study, specifically those not related to the causes of bee colony losses.
   - **Loss Data**: The dataset was refined to single out data representing the losses attributed to mites and climate change, allowing us to focus on these critical factors.

3. **Exploratory Focus**:
   - **State and Year Analysis**: By focusing on data that represented colony losses across different states and years, we aimed to explore the geographical and temporal impacts of these factors on bee colonies.

@@

@@

# Create a 'data' directory within the project
dir.create(here("data"))

@@

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Scatterplot of Value by State.ANSI
ggplot(data_filtered, aes(x = State.ANSI, y = Value, color = factor(Year))) +
  geom_point() +
  scale_y_continuous(labels = comma) +
  labs(title = "Scatterplot of Value by State.ANSI", x = "State.ANSI", y = "Value") +
  theme_minimal()

@@

# Print the plot
print(p)

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(reshape2)
library(cluster)
library(car)
library(factoextra)
library(rpart)
library(rpart.plot)
library(randomForest)
library(e1071)

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Descriptive Statistics
print("Descriptive Statistics:")
summary(data_clean)
str(data_clean)
```

@@

1. **Varroa Mites**: 
   - The data ranges from 8.00 to 67.20 with a median value of 26.80 and a mean of 32.41. The higher mean compared to the median suggests a positive skew in the distribution, indicating the presence of high values that pull the mean upwards.

2. **Other Pests**:
   - The values range from 1.70 to 42.30 with a median of 6.40 and a mean of 11.55. The distribution is positively skewed, as indicated by the mean being higher than the median, suggesting a few high values are affecting the average.

3. **Disease**:
   - The range is from 0.100 to 47.800, with a median of 1.000 and a mean of 6.718. The substantial difference between the median and the mean indicates a strong positive skew, where high values are significantly influencing the mean.

4. **Pesticides**:
   - The values range from 0.50 to 49.20, with a median of 5.70 and a mean of 10.34. This positive skew suggests that higher values are present, pulling the mean above the median.

5. **Other**:
   - The range is from 0.50 to 48.10, with a median of 3.40 and a mean of 10.51. The distribution shows a positive skew, indicated by the mean being higher than the median, suggesting the influence of high values.

6. **Unknown**:
   - The values range from 1.100 to 46.800, with a median of 4.400 and a mean of 8.718. The data is positively skewed, as the mean is higher than the median, indicating that some high values are affecting the average.

@@

ggplot(data_long, aes(x = state, y = percentage, fill = factor)) +
  geom_bar(stat = "identity") +
 theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Factors Affecting Bees by State", y = "Percentage", x = "State")
```
The bar chart represents various factors affecting bees across different states, with each factor color-coded. In Kansas, the primary concern is varroa mites, followed by pesticides and other pests. Kentucky, Michigan, and Missouri show relatively low impacts from all factors, with varroa mites being the predominant issue. Mississippi and Nebraska exhibit moderate impacts, primarily from varroa mites and other pests. Ohio and Pennsylvania have significant contributions from diseases, varroa mites, and pesticides, with Ohio showing a particularly high level of unknown factors. Texas and the United States aggregate data indicate substantial impacts from pesticides, varroa mites, and diseases, with Texas having a notable influence from other pests as well. Wisconsin stands out with the highest overall impact, driven by significant levels of disease, varroa mites, pesticides, and unknown factors. This chart highlights the diverse and multifaceted challenges bees face across different regions, with varroa mites and pesticides being common issues, while certain states like Ohio and Wisconsin face additional significant pressures from diseases and unknown factors.

@@

```
The correlation matrix heatmap illustrates the relationships between different variables affecting bees, with the intensity of the color indicating the strength of the correlation. Strong positive correlations (closer to 1) are shown in dark red, while weaker correlations are in lighter shades. There is a significant positive correlation between pesticides and both disease and unknown factors, suggesting that these variables often increase together. Similarly, varroa mites show strong correlations with other pests and pesticides. Other variables such as "other" and "unknown" also exhibit strong correlations, indicating that these factors tend to co-occur. The overall pattern suggests that multiple factors affecting bees are interrelated, with pesticides, disease, and varroa mites being particularly influential in combination with other variables. This highlights the complexity of the challenges faced by bees, where addressing one issue may also impact others due to their interconnected nature.

@@

write.csv(data_hcny_clean, here("hcnydata_cleaned.csv"), row.names = FALSE)

```{r}
# Load necessary library
library(e1071)
here()

@@

# If you want to visualize skewness using a bar plot:
skewness_df <- data.frame(Variable = names(skewness_results), Skewness = skewness_results)
ggplot(skewness_df, aes(x = Variable, y = Skewness)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Skewness of Variables", x = "Variables", y = "Skewness") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

```
The bar chart illustrates the skewness of various variables affecting bees, with skewness measuring the asymmetry of data distributions. The variable "disease" exhibits the highest skewness, slightly above 2.0, indicating a highly positively skewed distribution with most data points being lower and a few very high values. "Unknown" also shows a high skewness close to 2.0, reflecting a similar pattern of asymmetry. "Pesticides" have a skewness around 1.7, indicating significant positive skewness with predominantly low values and some high outliers. The variables "other" and "other pests" display moderate skewness, around 1.5, suggesting less extreme but noticeable asymmetry. In contrast, "varroa mites" exhibit the lowest skewness, just above 0.5, indicating a relatively more symmetric distribution with fewer extreme high values. Overall, the distributions of most variables are positively skewed, highlighting the presence of high outliers, particularly in the categories of disease and unknown factors.
```{r}

# Load necessary library
library(MASS)
library(ggplot2)
library(e1071)

@@

# Visualize skewness using a bar plot
skewness_df_transformed <- data.frame(Variable = names(skewness_results_transformed), Skewness = skewness_results_transformed)
ggplot(skewness_df_transformed, aes(x = Variable, y = Skewness)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Skewness of Variables after Transformation", x = "Variables", y = "Skewness") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
# Normality check
shapiro_results <- lapply(data_hcny_clean_transformed[-1], function(column) shapiro.test(column))
print("Shapiro-Wilk Test for Normality:")
print(shapiro_results)

```
The two sets of skewness values show the effect of data transformations on the distributions of various variables affecting bees. Initially, the skewness values were as follows:

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Cluster Analysis
data_cluster <- data_hcny_clean %>%
  filter(state != "United States") %>%
  dplyr::select(-state)

hc <- hclust(dist(data_cluster), method = "complete")
plot(hc, labels = data_hcny_clean$state[data_hcny_clean$state != "United States"], main = "Hierarchical Clustering of States")
```
The dendrogram represents the results of hierarchical clustering of states based on variables affecting bees. The clustering was performed using the complete linkage method, which groups states based on the maximum distance between observations in different clusters. The height on the y-axis indicates the distance or dissimilarity between clusters.

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Principal Component Analysis (PCA)
data_pca <- data_hcny_clean %>%
  filter(state != "United States") %>%
  dplyr::select(-state) %>%
  scale()
```

>>>>

@@

Group Ten is conducting a comprehensive analysis of historical data from
various public agencies to evaluate the impact of Varroa mites, American
Foulbrood, and global warming on hive losses in the United States.This
research utilizes extensive datasets from the National Agricultural
Statistics Service, the Agricultural Statistics Board, and the United
States Department of Agriculture (USDA), encompassing multiple years of
data. The data highlights the topic on hive losses attributed to mites,
bacterial infections, and environmental factors related to global
warming. Through advanced data visualization techniques in R, we aim to
demonstrate and validate the detrimental effects of these factors on
honeybee colonies, highlighting the consequent implications for honey
production and broader food security. ***Dennis and Kemp's study on
honey bee hive collapse [@hivecollapse] provides important insights into
Allee effects and ecological resilience.*** ***Warmer climate may impact
bees[@2024warmer]***

@@

Honeybees have served as nature's pollinators for centuries. With their
relationship with humans documented as far back as ancient Egyptian and
Hindu cultures. Historically, humans have maintained beehives and
utilized honey for its medicinal properties in various civilizations,
including the Egyptians, Assyrians, Chinese, Greeks, and Romans. Honey's
natural antibacterial qualities made it a valuable treatment for wounds.
This was a practice continued by Romans and Russians during World War I.
Honeybees and other pollinators are essential for food production and
nutritional security. But even though they face numerous survival
challenges Varroa mites pose a significant threat to bee colonies. These
tiny red-brown parasites live on adult honeybees and reproduce on larvae
and pupae. Another major threat is American Foulbrood Disease (AFB).
This is caused by the bacterium Paenibacillus larvae, which is fatal to
honeybee larvae and requires incineration of infected hives to prevent
its spread. The impact of climate change on honeybee colony losses is a
growing area of research. While correlations have been found between
higher winter temperatures and increased colony losses, the effects of
warmer autumn and winter temperatures on colony dynamics and survival
require further investigation. ***Dennis and Kemp's study on honey bee
hive collapse [@hivecollapse] provides important insights into Allee
effects and ecological resilience.*** [@usda_honey_bees;
@usda_nass]\*\*\*

***Warmer climate may impact bees[@2024warmer]***

@@

# Plot with ggplot2
ggplot(honey_data, aes(x = "", y = percentage, fill = component)) +
  geom_bar(width = 0.8, stat = "identity", color = "white") +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = honey_data$color) +
  geom_text(aes(x=1.8, label = paste0(percentage, "%")), position = position_stack(vjust = 0.6), size = 3, color = "black") +
  labs(
    title = "COMPOSITION OF HONEY",
    subtitle = "Illustration of honey components by percentage",
    fill = NULL
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5, family = "lobster"),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 12)
  )
```

### Composition of Honey

The first image is a pie chart that illustrates the composition of honey
by percentage. The main components are:

@@

### Spread of Varroa Mites

The second image is a map illustrating the spread of Varroa mites in the
United States from 1986 to 1995. The map uses different colors to
represent the year of the first appearance of Varroa mites in each
state:

@@

The map reveals the progressive spread of Varroa mites from the
mid-1980s to the mid-1990s, affecting bee colonies across the country.
The earliest appearances were concentrated in specific regions and
gradually spread to more states over time. This visual representation
highlights the increasing geographical distribution of Varroa mites and
underscores the growing challenge they pose to beekeeping and
agricultural industries in the United States. [@varroamap] ***Varroa Map
of infestation periods***

@@

### Detailed Description:

-   **Trend Analysis**:
    -   From 1900 to around 1975, the Earth's surface temperature shows
        considerable variability with some notable periods of cooling
        and warming.
    -   Post-1975, there is a distinct upward trend in temperature
        anomalies for all three measurements, indicating a general
        warming trend.
-   **Measurement Comparison**:
    -   The Earth's surface (blue line) shows the most extended
        historical data and fluctuates more in the earlier years
        compared to the troposphere measurements.
    -   The lower troposphere measurements (RSS in red and UAH in green)
        start around 1975 and generally follow the same upward trend as
        the surface measurements but with less variability.
    -   The RSS and UAH measurements align closely with each other
        post-2000, indicating consistent trends in the lower troposphere
        temperature anomalies.
-   **Anomalies & Peaks**:
    -   The Earth's surface shows significant peaks and troughs,
        especially in the early 20th century and around the year 2000.
    -   The lower troposphere measurements show peaks and align closely,
        particularly after the year 2000, reflecting similar patterns in
        temperature anomalies.

### Key Points:

-   The overall trend shows an increase in temperature anomalies from
    1900 to 2025.
-   Significant variability is seen in Earth's surface temperatures in
    the early years.
-   Post-1975, a clear warming trend is observed across all
    measurements.
-   Lower troposphere data (RSS and UAH) begins around 1975 and follows
    the Earth's surface trend with less variability.
-   Post-2000, the RSS and UAH data closely align, indicating consistent
    temperature trends in the lower troposphere.
-   There are noticeable peaks in the temperature anomalies around the
    year 2000 and later, reflecting periods of significant temperature
    increases.

### Impact on Bees, Varroa Mites, and Colony Loss:

-   **Bees**:
    -   **Thermal Stress**: Increased temperature anomalies can cause
        thermal stress on bees, affecting their ability to forage,
        navigate, and perform essential colony tasks.
    -   **Floral Resources**: Warming temperatures can alter the
        availability and distribution of floral resources, impacting bee
        nutrition and health.
    -   **Reproduction Cycles**: Changes in temperature can disrupt the
        reproductive cycles of bees, potentially leading to mismatches
        with the availability of pollen and nectar.
-   **Varroa Mites**:
    -   **Increased Reproduction**: Warmer temperatures can accelerate
        the reproduction rates of varroa mites, exacerbating
        infestations within bee colonies.
    -   **Extended Activity Period**: Higher temperatures can lengthen
        the active period of varroa mites throughout the year, leading
        to prolonged periods of stress on bee populations.
-   **Colony Loss**:
    -   **Health Decline**: Combined effects of thermal stress on bees
        and increased varroa mite infestations can lead to a decline in
        colony health.
    -   **Colony Collapse Disorder (CCD)**: Temperature anomalies can
        contribute to conditions that favor Colony Collapse Disorder,
        where the majority of worker bees disappear, leaving behind the
        queen and immature bees.
    -   **Mortality Rates**: Sustained temperature increases can lead to
        higher mortality rates within bee colonies, significantly
        impacting beekeeping and agricultural pollination services.

### Conclusion:

The chart highlights a clear trend of increasing temperature anomalies,
which can have profound implications on bee populations, varroa mite
dynamics, and overall colony health. Understanding these trends is
crucial for developing strategies to mitigate the adverse effects on
bees and ensure the sustainability of pollination services essential for
ecosystems and agriculture.

[@epa_climate] ***Global warming trends from 1900-2024***

@@

1.  **USDA National Agricultural Statistics Service (NASS)**:
    -   Provides comprehensive agricultural data, including statistics
        on honey production and colony health.
    -   Source: [USDA - National Agricultural Statistics Service -
        Surveys - honey bee surveys and
        reports](https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/).
2.  **Bee Informed Partnership**:
    -   Offers detailed insights and research on bee colony management
        and health, contributing valuable information on the status and
        trends of bee populations.
    -   Source: [Index Catalog // USDA Economics, Statistics and Market
        Information
        System](https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en).

@@

@@

@@

@Article{hivecollapse, author = {Brian Dennis and William P. Kemp},
title = {How Hives Collapse: Allee Effects, Ecological Resilience, and
the Honey Bee}, journal = {PLOS ONE}, year = {2016}, volume = {11},
number = {2}, pages = {e0150055}, doi = {10.1371/journal.pone.0150055},
url = {https://doi.org/10.1371/journal.pone.0150055} A study performed
to determine various elements impacting bee hive collapse}

@USDAWebsite{varroamap, author = {sdns6mchl4}, title = {Varroa mite
spread in the United States}, year = {2016}, month = {February 24}, url
=
{https://www.beesource.com/threads/varroa-mite-spread-in-the-united-states.365462/},
note = {Beesource Beekeeping Forums} Map highlighting states of Varroa
infestation from periods 1986-1995}

@USDAWebsite{usda_honey_bees, author = {{USDA Economics, Statistics and
Market Information System}}, title = {Index Catalog}, year = {n.d.-a},
url =
{https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en},
note = {Accessed: 2024-07-31} Data sets}

@USDAWebsite{usda_nass, author = {{USDA - National Agricultural
Statistics Service}}, title = {Surveys - honey bee surveys and reports},
year = {n.d.}, url =
{https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/},
note = {Accessed: 2024-07-31} Data sets}

@EPAWebsite{epa_climate, author = {{U.S. Environmental Protection
Agency}}, title = {Climate Change Indicators: U.S. and Global
Temperature}, year = {2024}, howpublished =
{\url{https://www.epa.gov/climate-indicators/climate-change-indicators-us-and-global-temperature}},
note = {Accessed: 2024-08-01} Climate change from periods of 1900-2024}

@article{2024warmer, title={Warmer autumns and winters could reduce
honey bee overwintering survival with potential risks for pollination
services}, author={Rajagopalan, K. and DeGrandi-Hoffman, G. and Pruett,
M. and others}, journal={Scientific Reports}, volume={14}, pages={5410},
year={2024}, publisher={Nature Publishing Group},
doi={10.1038/s41598-024-55327-8} Peer reviewed article concering global
warming possibly impacting bee hive survivability and viability during
winter periods}

@@

@@

The United States Department of Agriculture (USDA) website served as our
data source. We downloaded CSV files containing vital information on
honey production, colony health, and factors impacting bee populations.
By using USDA data, we ensured our analysis was based on reliable,
comprehensive, and relevant records widely recognized in agricultural
and ecological research.

@@

1.  **Removal of Irrelevant Data**:
    -   **Blank Spaces**: We eliminated any blank spaces within the
        datasets to ensure data integrity.
    -   **Non-Pertinent Columns**: Columns that were not directly
        related to our analysis objectives were removed to streamline
        the dataset.
2.  **Filtering Observations**:
    -   **Focus on Relevant Data**: We filtered out observations that
        did not directly pertain to our study, specifically those not
        related to the causes of bee colony losses.
    -   **Loss Data**: The dataset was refined to single out data
        representing the losses attributed to mites and climate change,
        allowing us to focus on these critical factors.
3.  **Exploratory Focus**:
    -   **State and Year Analysis**: By focusing on data that
        represented colony losses across different states and years, we
        aimed to explore the geographical and temporal impacts of these
        factors on bee colonies.

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Install necessary packages
if (!require("DiagrammeR")) install.packages("DiagrammeR")

# Create the flowchart using DiagrammeR
grViz("
  digraph Data_Analysis_Flowchart {
    graph [layout = dot, rankdir = LR, label = 'Workflow Schematic', fontsize = 75, labelloc = 't']

    node [style = filled, fontsize = 50, width = 6, height = 6, penwidth = 8]

    Define_Objectives -> Collect_Data
    Collect_Data -> Clean_Data
    Clean_Data -> Analyze_Data
    Analyze_Data -> Interpret_Results
    Interpret_Results -> Visualize_Data
    Visualize_Data -> Compile_Report
    Compile_Report -> Review_Revise
    Review_Revise -> Present_Report
  }
")

@@

here()
# Create a 'data' directory within the project
dir.create(here("data"))

@@

str(data_filtered)

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Scatterplot of Value by State.ANSI
ggplot(data_filtered, aes(x = State.ANSI, y = Value, color = factor(Year))) +
  geom_point() +
  labs(title = "Scatterplot of Value by State.ANSI", x = "State.ANSI", y = "Value") +
  theme_minimal()
```

@@

# Print the plot
print(p)
```

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(reshape2)
library(cluster)
library(car)
library(factoextra)
library(rpart)
library(rpart.plot)
library(randomForest)
library(e1071)
```

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Descriptive Statistics
print("Descriptive Statistics:")
summary(data_clean)

@@

1.  **Varroa Mites**:
    -   The data ranges from 8.00 to 67.20 with a median value of 26.80
        and a mean of 32.41. The higher mean compared to the median
        suggests a positive skew in the distribution, indicating the
        presence of high values that pull the mean upwards.
2.  **Other Pests**:
    -   The values range from 1.70 to 42.30 with a median of 6.40 and a
        mean of 11.55. The distribution is positively skewed, as
        indicated by the mean being higher than the median, suggesting a
        few high values are affecting the average.
3.  **Disease**:
    -   The range is from 0.100 to 47.800, with a median of 1.000 and a
        mean of 6.718. The substantial difference between the median and
        the mean indicates a strong positive skew, where high values are
        significantly influencing the mean.
4.  **Pesticides**:
    -   The values range from 0.50 to 49.20, with a median of 5.70 and a
        mean of 10.34. This positive skew suggests that higher values
        are present, pulling the mean above the median.
5.  **Other**:
    -   The range is from 0.50 to 48.10, with a median of 3.40 and a
        mean of 10.51. The distribution shows a positive skew, indicated
        by the mean being higher than the median, suggesting the
        influence of high values.
6.  **Unknown**:
    -   The values range from 1.100 to 46.800, with a median of 4.400
        and a mean of 8.718. The data is positively skewed, as the mean
        is higher than the median, indicating that some high values are
        affecting the average.

@@

ggplot(data_long, aes(x = state, y = percentage, fill = factor)) +
  geom_bar(stat = "identity") +
 theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Factors Affecting Bees by State", y = "Percentage", x = "State")
```

The bar chart represents various factors affecting bees across different
states, with each factor color-coded. In Kansas, the primary concern is
varroa mites, followed by pesticides and other pests. Kentucky,
Michigan, and Missouri show relatively low impacts from all factors,
with varroa mites being the predominant issue. Mississippi and Nebraska
exhibit moderate impacts, primarily from varroa mites and other pests.
Ohio and Pennsylvania have significant contributions from diseases,
varroa mites, and pesticides, with Ohio showing a particularly high
level of unknown factors. Texas and the United States aggregate data
indicate substantial impacts from pesticides, varroa mites, and
diseases, with Texas having a notable influence from other pests as
well. Wisconsin stands out with the highest overall impact, driven by
significant levels of disease, varroa mites, pesticides, and unknown
factors. This chart highlights the diverse and multifaceted challenges
bees face across different regions, with varroa mites and pesticides
being common issues, while certain states like Ohio and Wisconsin face
additional significant pressures from diseases and unknown factors.

@@

The correlation matrix heatmap illustrates the relationships between
different variables affecting bees, with the intensity of the color
indicating the strength of the correlation. Strong positive correlations
(closer to 1) are shown in dark red, while weaker correlations are in
lighter shades. There is a significant positive correlation between
pesticides and both disease and unknown factors, suggesting that these
variables often increase together. Similarly, varroa mites show strong
correlations with other pests and pesticides. Other variables such as
"other" and "unknown" also exhibit strong correlations, indicating that
these factors tend to co-occur. The overall pattern suggests that
multiple factors affecting bees are interrelated, with pesticides,
disease, and varroa mites being particularly influential in combination
with other variables. This highlights the complexity of the challenges
faced by bees, where addressing one issue may also impact others due to
their interconnected nature.

@@

write.csv(data_hcny_clean, here("hcnydata_cleaned.csv"), row.names = FALSE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load necessary library
library(e1071)
here()

@@

# If you want to visualize skewness using a bar plot:
skewness_df <- data.frame(Variable = names(skewness_results), Skewness = skewness_results)
ggplot(skewness_df, aes(x = Variable, y = Skewness)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Skewness of Variables", x = "Variables", y = "Skewness") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

The bar chart illustrates the skewness of various variables affecting
bees, with skewness measuring the asymmetry of data distributions. The
variable "disease" exhibits the highest skewness, slightly above 2.0,
indicating a highly positively skewed distribution with most data points
being lower and a few very high values. "Unknown" also shows a high
skewness close to 2.0, reflecting a similar pattern of asymmetry.
"Pesticides" have a skewness around 1.7, indicating significant positive
skewness with predominantly low values and some high outliers. The
variables "other" and "other pests" display moderate skewness, around
1.5, suggesting less extreme but noticeable asymmetry. In contrast,
"varroa mites" exhibit the lowest skewness, just above 0.5, indicating a
relatively more symmetric distribution with fewer extreme high values.
Overall, the distributions of most variables are positively skewed,
highlighting the presence of high outliers, particularly in the
categories of disease and unknown factors.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load necessary library
library(MASS)
library(ggplot2)
library(e1071)

@@

# Visualize skewness using a bar plot
skewness_df_transformed <- data.frame(Variable = names(skewness_results_transformed), Skewness = skewness_results_transformed)
ggplot(skewness_df_transformed, aes(x = Variable, y = Skewness)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Skewness of Variables after Transformation", x = "Variables", y = "Skewness") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
# Normality check
shapiro_results <- lapply(data_hcny_clean_transformed[-1], function(column) shapiro.test(column))
print("Shapiro-Wilk Test for Normality:")
print(shapiro_results)
```

The two sets of skewness values show the effect of data transformations
on the distributions of various variables affecting bees. Initially, the
skewness values were as follows:

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Cluster Analysis
data_cluster <- data_clean %>%
  filter(state != "United States") %>%
  dplyr::select(-state)

hc <- hclust(dist(data_cluster), method = "complete")
plot(hc, labels = data_clean$state[data_clean$state != "United States"], main = "Hierarchical Clustering of States")
```

The dendrogram represents the results of hierarchical clustering of
states based on variables affecting bees. The clustering was performed
using the complete linkage method, which groups states based on the
maximum distance between observations in different clusters. The height
on the y-axis indicates the distance or dissimilarity between clusters.

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Principal Component Analysis (PCA)
data_pca <- data_clean %>%
  filter(state != "United States") %>%
  dplyr::select(-state) %>%
  scale()
```
